\LARGE \bf Fast Approximate 2D Ray Casting for Accelerated Localization

### 2017-03-02 paper/main.tex
added submission version of paper

% The following packages can be found on http:\\www.ctan.org
\usepackage{graphicx}
\usepackage{tabu}
\usepackage{float}
% \usepackage{graphics} % for pdf, bitmapped graphics files
%\usepackage{epsfig} % for postscript graphics files
%\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
\usepackage{amsmath} % assumes amsmath package installed
%\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{tikz}
\usetikzlibrary{matrix}
\usetikzlibrary{shapes.geometric, arrows}
\usepackage{algorithm}
% \usepackage[]{algorithm2e}
\usepackage{algorithmic}

\newcommand{\img}[1]{\begin{center}\includegraphics[width=\columnwidth]{{#1}}\end{center}}

\newcommand{\imgFrame}[1]{\begin{center}\frame{\includegraphics[width=\columnwidth]{{#1}}}\end{center}}
\newcommand{\INDSTATE}[1][1]{\STATE\hspace{#1\algorithmicindent}}

\newcommand{\gridScale}{1.25}

%: Margin note commands
\usepackage[usenames,dvipsnames]{xcolor}
\setlength{\marginparwidth}{0.5in}
\let\oldmarginpar\marginpar
\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%
{\raggedright\footnotesize #1}}
\newcommand{\skcomment}[2]{{\color{red}#1} \marginpar{\color{blue}\tiny #2 (SK)}}

% \begin{center}\frame{\includegraphics[width=\columnwidth]{synthetic.png}}\end{center}

@@

Lorem Ipsum is simply dummy text of the printing and typesetting industry. Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book. It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged. It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.

@@

Several well known algorithms exist for ray casting in two dimensional spaces such as Bresenham's Line algorithm \cite{bresenham} and ray marching \cite{raymarching}. Both algorithms work by iteratively checking points in the map starting at the query point and moving in the ray direction until an obstacle is discovered. This process may require hundreds of memory reads per ray cast depending on the distance to the nearest obstacle, and does not provide constant time performance.

To combat the computational challenges of ray casting while localizing in a two-dimensional map, Thrun et al. \cite{localization} suggest the use of a large three-dimensional lookup table (LUT) to store expected ranges for each discrete $(x,y,\theta)$ state. While this is simple to implement and does result in large speed improvements as compared to ray casting, it can be prohibitively memory intensive for large maps and/or resource constrained systems. In a 2000 by 2000 occupancy map, storing ranges for 200 discrete directions would require over 1.5GB. While this memory requirement may be acceptable in many cases, it scales with the square of map size - a 4000 by 4000 map would require over 6GB for the same angular discretization.

We propose an analogous data structure to the three-dimensional LUT called a Compressed Directional Distance Transform (CDDT) which requires significantly less memory and precomputation time while still allowing near constant time ray casting queries. The contributions of this work are the following. Firstly, we develop a problem formulation and provide details and analysis of our method. Secondly, we implement the proposed algorithm and compare its performance to alternate two-dimensional ray casting methods. Finally, we discuss potential extensions to the CDDT algorithm which could be the basis of future research. We observe two orders of magnitude improvement in memory requirements, with little sacrifice in computation time, when compared to the lookup table methods. Additionally, we observe a large speedup when compared to the other ray casting methods considered with similar memory requirements.

@@

Bresenham's line algorithm \cite{bresenham} incrementally determines the set of pixels that approximate the trajectory of query ray starting from the query point $(x,y)_{query}$ and progressing in the $\theta_{query}$ direction one pixel at a time. The algorithm terminates once the nearest occupied pixel is discovered, and the euclidean distance between that occupied pixel and $(x,y)_{query}$ is reported. This algorithm is widely implemented in particle filters due to its simplicity and ability to operate on a dynamic map. The primary disadvantage is that it is slow, potentially requiring hundreds of memory accesses for a single ray cast. While actual performance is highly environment dependent, Bresenham’s Line algorithm is linear in map size in the worst case.

@@

As previously described, a common acceleration technique for two dimensional ray casting is to precompute the ray distances for every state in a discrete grid and store the results in a three dimensional LUT. Any ray casting technique may be used to compute the table. In the authors’ implementation ray casting is used to populate the LUT since it has low initialization time and is significantly faster than Bresenham’s Line. Theoretically the LUT approach has constant query runtime, though actual performance is dependent on access patterns as CPU caching effects are significant in practice.

State space discretization implies approximate results, since intermediate states must be rounded onto the discrete grid. While the effect of rounding query position is small, rounding $\theta_{query}$ may have significant effects. As the ray moves further away from the query point along the theta direction, angular discretization error accumulates. For queries $(x,y,\theta)_{query}$ discretized into $\lfloor(x,y,\theta)\rceil$, the distance between the end of the ray $\lfloor(x,y,\theta)\rceil$ and its projection onto the line implied by $(x,y,\theta)_{query}$ becomes large as the length of the ray increases. Although the error induced by discretization may be unacceptable for computer graphics applications, it is generally acceptable for probabilistic algorithms such as the particle filter since error is expected and directly modeled.

@@

Rather than simply rounding to the nearest grid state, one may improve accuracy of queries which lie between discrete states by querying the neighboring states and interpolating the results. Since we are evaluating the performance of these methods for use in a particle filter, we do not perform interpolation since the increased computation reduces the number of particles that may be maintained in real time, thereby effectively negating any potential increase in accuracy with respect to the stored map.

@@

\begin{figure}[h]
\img{coord_system.png}
\caption{Occupancy grid map coordinate system}
\label{coordinate_system}
\end{figure}

@@

Although the three dimensional table used to store precomputed ray cast solutions for a discrete state space is inherently large, it is highly compressible. This is most apparent in the cardinal directions, in which adjacent values along a particular dimension of the table increase by exactly one unit of distance for unobstructed positions as in Fig. \ref{cardinal_example}. Our data structure is designed to compress this redundancy while still allowing for fast queries in near constant time. We accomplish this though through what we refer to as a Compressed Directional Distance Transform described here.

\begin{figure}[h]
\begin{center}
\begin{tikzpicture}
\fill[black,scale=\gridScale] (0.5,0.5) rectangle (1,1);
\fill[black,scale=\gridScale] (-0.5,0) rectangle (0,0.5);
\fill[black,scale=\gridScale] (0,-0.5) rectangle (0.5,0);
\fill[black,scale=\gridScale] (0.5,-1) rectangle (1,-0.5);
\draw[scale=\gridScale,step=0.5cm,color=white] (-1,-1.1) grid (1,1.1);
\draw[scale=\gridScale,step=0.5cm,color=gray] (-1,-1) grid (1,1);
\end{tikzpicture}
\begin{tikzpicture}
\draw[scale=\gridScale,step=0.5cm,color=white] (-1,-1.1) grid (1,1.1);
\draw[scale=\gridScale,step=0.5cm,color=gray] (-1,-1) grid (1,1);
\matrix[matrix of nodes,nodes={inner sep=0pt,text width=.5cm,align=center,minimum height=0.45cm, scale=\gridScale}]{
3 & 2 & 1 & 0 \\
1 & 0 & - & - \\
2 & 1 & 0 & - \\    
3 & 2 & 1 & 0\\};
\end{tikzpicture}
\end{center}
\caption{Occupancy grid (left) and associated LUT for $\theta=0$ (right).}
\label{cardinal_example}
\end{figure}

The euclidean distance transform of an occupancy grid map stores the distance to the nearest obstacle in any direction $\theta$ for each possible $\lfloor(x,y)\rceil$ in the map. In contrast, a directional distance transform (DDT) stores the distance to the nearest obstacle in a particular direction $\lfloor\theta\rceil$ for every $\lfloor(x,y)\rceil$. The key difference between a single $\theta$ slice of the LUT and a directional distance transform for the same $\theta$ is the way in which it is computed and indexed. To compute the LUT slice, ray casting is performed in the $\theta$ direction for every $\lfloor(x,y)\rceil$. At runtime, each $(x,y)_{query}$ is discretized to $\lfloor(x,y)_{query}\rceil$ and the LUT slice is directly indexed.

\begin{figure}[h]
\img{ddt_lut_slice_comparison_map.png}
\caption{Comparison between a DDT and a LUT slice for the same value of theta. Each row in the DDT is characterized by a sawtooth function.}
\label{ddt_lut_comparison}
\end{figure}

In contrast, to compute the DDT, the obstacles in the map are rotated about the origin by $-\lfloor\theta\rceil$ and ray casting is implicitly performed in the $\theta=0$ direction, as demonstrated by Fig. \ref{ddt_lut_comparison}. Then, to index the table for a query $(x,y,\theta)_{query}$ one discretizes to $\lfloor(x,y,\theta)_{query}\rceil$, rotates $\lfloor(x,y)_{query}\rceil$ about the origin, and uses the rotated coordinates $\lfloor(x,y,\theta)_{query}\rceil_{rot}$ to (implicitly) index into the DDT. Thus, roughly the same operation is computed in both the DDT and the LUT, but while one changes the ray cast direction to populate the LUT, one rotates scene geometry and ray casts in a constant direction to populate the DDT.

@@

\begin{figure}[h]
\img{sawtooth_to_cddt.png}
\caption{Demonstration of compression from a sawtooth function to a list of zero points. Consecutively valued zero points exist where walls are greater than one pixel thick.}
\label{sawtooth_to_cddt}
\end{figure}

The conversion from a DDT to a CDDT slice is performed by storing the $x$ coordinate of each zero point in the coordinate space of the DDT for every discrete row as shown in Fig. \ref{sawtooth_to_cddt}. At query time, the sawtooth function (i.e. exactly the distance to the nearest obstacle in the query direction) encoded in the DDT may be quickly recovered by performing binary search for nearest zero point in the correct row of the CDDT slice. We refer to the list of zero points for a single row as a CDDT bin. Similar to the LUT, the full CDDT is defined as every CDDT slice for all discrete values of $\theta$.

For performance, it is not necessary to compute the full DDT, but rather the CDDT may be directly constructed by projecting each obstacle into the coordinate space of the DDT for every $\lfloor\theta\rceil$ and sorting its $x$ coordinate in the CDDT bin corresponding to its $y$ coordinate and $\lfloor\theta\rceil$. After all geometry has been projected into the CDDT, each bin is sorted to facilitate later indexing. Not only does the direct construction of the CDDT greatly reduce the amount of memory required to store the LUT, it also reduces data structure precomputation time since ray casting is not necessary.

While conceptually simple, implementing the CDDT data structure construction and traversal routines requires careful consideration in order to capture all edge cases and to minimize unnecessary computation. In this sense, it is more complex to implement than the alternatives considered, however there are many opportunities for optimization which yield real-world speed up. To ease this burden, we provide our implementation as well as Python wrappers with an Apache 2.0 license.

@@

A side effect of extracting zero points from each row of the DDT is the introduction of rotational symmetry. Not only can a row of the DDT in the $\theta$ direction be reconstructed from the zero points, but also a row of the DDT in the $\theta+\pi$ direction as in Fig. \ref{bidirectional_zero_points}. Therefore, one only need compute CDDT slices for $0\leq\lfloor\theta\rceil<\pi$ and the DDT for $0\leq\lfloor\theta\rceil<2\pi$ may be inferred, resulting in a factor of further two reduction in memory usage.

@@

% \subsection{Incremental Update}

While the authors did not implement the necessary changes, we conjecture that the CDDT algorithm could be modified to allow incremental updates to the map. In both ray marching and the LUT approach, any change in the map introduces inconsistencies to the acceleration data structure requiring a full re-initialization. As demonstrated by our experiments, initializing the three dimensional LUT is a highly computational process. Similarly, computing a euclidean distance transform for ray marching can be expensive. Therefore neither ray casting nor the LUT can be used with dynamic maps such as in SLAM algorithms. Despite being the slowest option evaluated, Bresenham's Line algorithm is one of the most commonly used techniques since it operates directly on the grid map without an acceleration data structure.

@@

% \subsection{Pruning}
\begin{figure}[h]
\begin{center}\includegraphics[width=6.5cm]{edge_map.png}\end{center}
\caption{Example map left, morphological edge map right.}
\label{edge_map}
\end{figure}

By removing entries in the CDDT which can never possibly result in a ray collision, it is possible to further reduce the memory footprint of the data structure. Consider a 3x3 block of obstacles in an otherwise empty map as in Fig. \ref{edge_map}. The center obstacle will never be the nearest neighbor in any ray casting query, because any such query would first intersect with one of the 8 surrounding obstacles. To exploit this, one can take the morphological edge map of the occupancy grid map prior to CDDT generation without loss of generality. Prior to performing ray casting on the CDDT data structure, we check the occupancy grid to ensure that the query point itself is not an obstacle and return a distance of zero if so in order to prevent incorrect results when ray casting from the middle of obstacles removed during the morphological operation. In dense maps, this results in a significant reduction in both memory usage and construction time.

% \begin{figure}[h]
% \img{slice0_pcddt_basement.png}
% \caption{A slice of the PCDDT visualized as the implied DDT. Blue line demonstrates the num}
% \label{pcddt:slice0}
% \end{figure}

Additionally, consider a line of obstacles aligned along the X-axis. Every element in this line will be projected into a single zero point bin in the $\theta=0$ CDDT slice. However, the middle elements of the line will never result in a collision. Any ray cast from points on the line of obstacles will return early in the occupancy grid check, and any ray cast from non-overlapping points co-linear in the $\theta$ or $\theta+\pi$ directions will intersect one of the edge-most obstacles in the line. Therefore in the $\theta=0$ CDDT slice, one may discard the zero points corresponding to the middle elements without introducing error. This form of optimization is simple to compute in the cardinal directions, but non-trivial for arbitrary $\lfloor\theta\rceil$ not aligned with an axis. Rather than analytically determining which obstacles may be discarded, it is simpler to prune the data structure by ray casting from every possible state, discarding any unused zero point.

Pruning does increase pre-computation time, rendering it incompatible with incremental update. However, the reduction of memory usage is worthwhile for static maps. In addition to memory reduction, we find that pruning slightly improves runtime performance, likely as a result of improved caching characteristics and the reduced number of zero points.

@@

% \skcomment{}{Do you think you can have a subsection called ``analysis'' which has a few propositions/theorems that say things like Memory requirement is $O(NM)$ and running time is $O(1)$? This may be hard given the deadline, but osmething to think about. }

The memory usage of the CDDT data structure is $O(n*\theta_{discretization}+|map|)$ where $n$ is the number of occupied pixels in the edge map. The occupancy grid map must be stored to check for overlaps between $(x,y)_{query}$ and obstacles prior to searching CDDT bins. For each edge pixel, a total of $\theta_{discretization}$ float values are stored in the CDDT bins. Due to the fact that we must sort each bin after CDDT construction, pre-computation time is at worst $O(n*\theta_{discretization} + S^2*\theta_{discretization}*log(S))$ for the same definition of n, where $S=norm(\text{map width},\text{map height})$. In practice it is closer to $O(n*\theta_{discretization} + S*\theta_{discretization})$ since each CDDT has a small number of elements on average, as evidenced by the high demonstrated compression ratio.

@@

The ray cast procedure has three general steps: projection into CDDT coordinate space, the search for nearby zero points, and the computation of distance given the nearest zero point. The first and last steps are simple arithmetic, and therefore are theoretically constant time. The second step involves searching for a value in a sorted array of zero points. As previously discussed in \ref{optimizations}, the number of zero points in each CDDT bin tends to be small and is bounded in map size. Thus, at worst the search operation requires $\log{(norm(\text{map width},\text{map height}))}$ which is a constant value for a fixed map. Therefore, for a given map, our algorithm provides constant time performance.

@@

We have implemented the proposed algorithm in the C++ programming language, as well as Bresenham’s Line, ray marching, and the LUT approach for comparison. Our source code is available for use and analysis [TODO add a link to github], and Python wrappers are also provided for easier usage. All benchmarks were performed on a desktop computer on a Intel Core i5-4590 CPU @ 3.30GHz with 16GB DDR3 ram @ 1333MHz, running Ubuntu 14.04.

>>>>

% The following packages can be found on http:\\www.ctan.org
\usepackage{graphicx}
\usepackage{tabu}
\usepackage{float}
% \usepackage{graphics} % for pdf, bitmapped graphics files
%\usepackage{epsfig} % for postscript graphics files
%\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
\usepackage{amsmath} % assumes amsmath package installed
%\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{tikz}
\usetikzlibrary{matrix}
\usetikzlibrary{shapes.geometric, arrows}
\usepackage{algorithm}
\usepackage{algorithmic}

\newcommand{\img}[1]{\begin{center}\includegraphics[width=\columnwidth]{{#1}}\end{center}}
\newcommand{\imgFrame}[1]{\begin{center}\frame{\includegraphics[width=\columnwidth]{{#1}}}\end{center}}
\newcommand{\INDSTATE}[1][1]{\STATE\hspace{#1\algorithmicindent}}
\newcommand{\gridScale}{1.25}

%: Margin note commands
% \usepackage[usenames,dvipsnames]{xcolor}
% \setlength{\marginparwidth}{0.5in}
% \let\oldmarginpar\marginpar
% \renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%
% {\raggedright\footnotesize #1}}
% \newcommand{\skcomment}[2]{{\color{red}#1} \marginpar{\color{blue}\tiny #2 (SK)}}

@@

Localization is an essential component for autonomous robots. A well-established localization approach combines ray casting with a particle filter, leading to a computationally expensive algorithm that is difficult to run on resource-constrained mobile robots. We present a novel data structure called the Compressed Directional Distance Transform for accelerating ray casting in two dimensional occupancy grid maps. Our approach has near constant time ray casting performance for a fixed size map, in contrast with other methods which exhibit long tail query time distributions. Our experimental results show that the proposed algorithm approximates the performance characteristics of reading from a three dimensional lookup table of ray cast solutions while requiring two orders of magnitude less memory and precomputation.

% Ray casting is a computationally expensive and commonly used operation in particle filter localization algorithms onboard mobile robots. We present a novel data structure called the Compressed Directional Distance Transform for accelerating ray casting in two dimensional occupancy grid maps. Our approach has near constant time ray casting performance for a fixed size map, in contrast with other methods which exhibit long tail query time distributions. Our experimental results show that the proposed algorithm approximates the performance characteristics of reading from a three dimensional lookup table of ray cast solutions while requiring two orders of magnitude less memory and precomputation.

@@

Several well known algorithms exist for ray casting in two dimensional spaces such as Bresenham's Line algorithm \cite{bresenham} and ray marching \cite{raymarching}. Both algorithms work by iteratively checking points in the map starting at the query point and moving in the ray direction until an obstacle is discovered. This process does not provide constant performance because the number of memory reads depends on the distance to the nearest obstacle.
% This process requires many memory reads, and does not provide constant time performance.

To combat the computational challenges of ray casting while localizing in a two-dimensional map, Fox et al. \cite{localization} suggest the use of a large three-dimensional lookup table (LUT) to store expected ranges for each discrete $(x,y,\theta)$ state. While this is simple to implement and does result in large speed improvements as compared to ray casting, it can be prohibitively memory intensive for large maps and/or resource constrained systems. In a 2000 by 2000 occupancy map, storing ranges for 200 discrete directions would require over 1.5GB. While this memory requirement may be acceptable in many cases, it scales with the square of map size - a 4000 by 4000 map would require over 6GB for the same angular discretization.

We propose an analogous data structure to the three-dimensional LUT called a Compressed Directional Distance Transform (CDDT) which requires significantly less memory and precomputation time while still allowing near constant time ray casting queries. The contributions of this work are the following. Firstly, we develop a problem formulation and provide details and analysis of our method. Secondly, we implement the proposed algorithm and compare its performance to alternate two-dimensional ray casting methods. Finally, we discuss potential extensions to the CDDT algorithm which could be the basis of future research.

We observe two orders of magnitude improvement in memory requirements, with little sacrifice in computation time, when compared to the lookup table methods. Additionally, we observe a large speedup when compared to the other ray casting methods considered with similar memory requirements.

@@

Bresenham's line algorithm \cite{bresenham} incrementally determines the set of pixels that approximate the trajectory of a query ray starting from the query point $(x,y)_{query}$ and progressing in the $\theta_{query}$ direction one pixel at a time. The algorithm terminates once the nearest occupied pixel is discovered, and the euclidean distance between that occupied pixel and $(x,y)_{query}$ is reported. This algorithm is widely implemented in particle filters due to its simplicity and ability to operate on a dynamic map. The primary disadvantage is that it is slow, potentially requiring hundreds of memory accesses for a single ray cast. While average performance is highly environment dependent, Bresenham’s Line algorithm is linear in map size in the worst case.

@@

As previously described, a common acceleration technique for two dimensional ray casting is to precompute the ray distances for every state in a discrete grid and store the results in a three dimensional LUT for later reference. Any ray casting technique may be used to compute the table. In the authors’ implementation ray marching is used for its fast performance compared to Bresenham's Line and its high accuracy. Theoretically the LUT approach has constant query runtime, though actual performance is dependent on access patterns as CPU caching effects are significant in practice.

State space discretization implies approximate results, since intermediate states must be rounded onto the discrete grid. While the effect of rounding query position is small, rounding $\theta_{query}$ may have significant effects since angular roundoff error accumulates with ray length. For queries $(x,y,\theta)_{query}$ discretized into $\lfloor(x,y,\theta)\rceil$, the distance between the end of the ray $\lfloor(x,y,\theta)\rceil$ and its projection onto the line implied by $(x,y,\theta)_{query}$ becomes large as the length of the ray increases. Although the error induced by discretization may be unacceptable for computer graphics applications, it is generally acceptable for probabilistic algorithms such as the particle filter since error is anticipated and directly modeled.

@@

Rather than simply rounding to the nearest grid state, one may improve accuracy of queries which lie between discrete states by querying the neighboring states and interpolating the results. Interpolation requires greater than one memory read and a small amount of arithmetic, and is therefore slower on average than rounding to the nearest grid state. Since we are evaluating the performance of these methods in the context of particle filter localization, the number of particles maintained is more important than ray cast accuracy so we do not perform interpolation.

@@

\begin{figure}[h!]
% \img{}
\begin{center}\includegraphics[width=7.6cm]{coord_system.png}\end{center}
\caption{Occupancy grid map coordinate system}
\label{coordinate_system}
\end{figure}

@@

Although the three dimensional table used to store precomputed ray cast solutions in a discrete state space is inherently large, it is highly compressible. This is most apparent in the cardinal directions, in which adjacent values along a particular dimension of the table increase by exactly one unit of distance for unobstructed positions as in Fig. \ref{cardinal_example}. Our data structure is designed to compress this redundancy while still allowing for fast queries in near constant time. We accomplish this though through what we refer to as a Compressed Directional Distance Transform described here.

\begin{figure}[h]
\begin{center}
\begin{tikzpicture}
\fill[black,scale=\gridScale] (0.5,0.5) rectangle (1,1);
\fill[black,scale=\gridScale] (-0.5,0) rectangle (0,0.5);
\fill[black,scale=\gridScale] (0,-0.5) rectangle (0.5,0);
\fill[black,scale=\gridScale] (0.5,-1) rectangle (1,-0.5);
\draw[scale=\gridScale,step=0.5cm,color=white] (-1,-1.1) grid (1,1.1);
\draw[scale=\gridScale,step=0.5cm,color=gray] (-1,-1) grid (1,1);
\end{tikzpicture}
\begin{tikzpicture}
\draw[scale=\gridScale,step=0.5cm,color=white] (-1,-1.1) grid (1,1.1);
\draw[scale=\gridScale,step=0.5cm,color=gray] (-1,-1) grid (1,1);
\matrix[matrix of nodes,nodes={inner sep=0pt,text width=.5cm,align=center,minimum height=0.45cm, scale=\gridScale}]{
3 & 2 & 1 & 0 \\
1 & 0 & - & - \\
2 & 1 & 0 & - \\    
3 & 2 & 1 & 0\\};
\end{tikzpicture}
\end{center}
\caption{Occupancy grid (left) and associated LUT slice for $\theta=0$ (right).}
\label{cardinal_example}
\end{figure}

The euclidean distance transform of an occupancy grid map stores the distance to the nearest obstacle in any direction $\theta$ for each possible $\lfloor(x,y)\rceil$ in the map. In contrast, what we call a directional distance transform (DDT) stores the distance to the nearest obstacle in a particular direction $\lfloor\theta\rceil$ for every $\lfloor(x,y)\rceil$. The key difference between a single $\theta$ slice of the LUT and a directional distance transform for the same $\theta$ is the way in which it is computed and indexed. To compute the LUT slice, ray casting is performed in the $\theta$ direction for every $\lfloor(x,y)\rceil$. At runtime, each $(x,y)_{query}$ is discretized to $\lfloor(x,y)_{query}\rceil$ and the LUT slice is directly indexed.

\begin{figure}[h]
\vspace{4pt}
\img{ddt_lut_slice_comparison_map.png}
\caption{Comparison between a DDT and a LUT slice for the same value of theta. Each row in the DDT is characterized by a sawtooth function.}
\label{ddt_lut_comparison}
\end{figure}

In contrast, to compute the DDT, the obstacles in the map are rotated about the origin by $-\lfloor\theta\rceil$ and ray casting is implicitly performed in the $\theta=0$ direction, as demonstrated by Fig. \ref{ddt_lut_comparison}. Then, to index the table for a query $(x,y,\theta)_{query}$, one discretizes to $\lfloor(x,y,\theta)_{query}\rceil$, rotates $\lfloor(x,y)_{query}\rceil$ about the origin, and uses the rotated coordinates $\lfloor(x,y,\theta)_{query}\rceil_{rot}$ to (implicitly) index into the DDT. Thus, roughly the same operation is computed in both the DDT and the LUT, but while one changes the ray cast direction to populate the LUT, one rotates scene geometry and ray casts in a constant direction to populate the DDT.

@@

\begin{figure}[h]
\img{sawtooth_to_cddt.png}
\caption{Demonstration of compression from a sawtooth function to a list of zero points. Consecutively valued zero points exist projected scene geometry spans greater than one unit along the x axis.}
\label{sawtooth_to_cddt}
\end{figure}

% The conversion from a DDT to a CDDT slice is done by storing the $x$ coordinate of each zero point in the coordinate space of the DDT for every discrete row as shown in Fig. \ref{sawtooth_to_cddt}.

The conversion from DDT to CDDT slice is done by storing the $x$ coordinates of each zero point for every row of the DDT as demonstrated for a single row in Fig. \ref{sawtooth_to_cddt}. At query time, the sawtooth function (i.e. exactly the distance to the nearest obstacle in the query direction) encoded in the DDT may be quickly recovered by performing binary search for nearest zero point in the correct row of the CDDT slice. We refer to the list of zero points for a single row as a CDDT bin. Similar to the LUT, the full CDDT is defined as every CDDT slice for all discrete values of $\theta$.

For performance, it is not necessary to compute or store the full DDT, but rather the CDDT may be directly constructed by projecting each obstacle into the coordinate space of the DDT for every $\lfloor\theta\rceil$ and storing its projected $x$ coordinate in the CDDT bin corresponding to its $y$ coordinate and $\lfloor\theta\rceil$. After all geometry has been projected into the CDDT, each bin is sorted to facilitate later indexing. Not only does the direct construction of the CDDT greatly reduce the amount of memory required to store a lookup table, it also reduces data structure precomputation time since ray casting every discrete state is not necessary.

While conceptually simple, implementing the CDDT data structure construction and traversal routines requires careful consideration in order to capture all edge cases and to minimize unnecessary computation. In this sense, it is more complex to implement than the alternatives considered, however there are many opportunities for optimization which yield real-world speed up. To ease this burden, we provide our implementation\footnote{https://github.com/kctess5/range\_libc} as well as Python wrappers under the Apache 2.0 license.

@@

A side effect of extracting zero points from each row of the DDT is the introduction of rotational symmetry. Not only can a row of the DDT in the $\theta$ direction be reconstructed from the zero points, but also a row of the DDT in the $\theta+\pi$ direction as in Fig. \ref{bidirectional_zero_points}. Therefore, one only need compute and store CDDT slices for $0\leq\lfloor\theta\rceil<\pi$ and the DDT for $0\leq\lfloor\theta\rceil<2\pi$ may be inferred by reversing the search direction, resulting in a factor of two reduction in memory usage.

@@

With small modifications, another factor of two reduction in CDDT size could be attained by storing zero points as 16 bit integers rather than 32 bit floats with little precision loss.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Incremental Update %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

While the authors did not implement the necessary changes, we conjecture that the CDDT algorithm could be modified to allow incremental updates to the map. In both ray marching and the LUT approach, any change in the map introduces inconsistencies to the acceleration data structure requiring a full re-initialization. As demonstrated by our experiments, initializing the three dimensional LUT is a highly computational process. Similarly, computing a euclidean distance transform for ray marching can be expensive. Therefore neither ray marching nor the LUT can be easily used with dynamic maps such as in SLAM algorithms. Despite being the slowest option evaluated, Bresenham's Line algorithm is one of the most commonly used techniques since it operates directly on the grid map without an acceleration data structure.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Pruning %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[h]
\begin{center}\includegraphics[width=6.5cm]{edge_map.png}\end{center}
\caption{Example map left, morphological edge map right.}
\label{edge_map}
\end{figure}

By removing entries in the CDDT which can never possibly result in a ray collision, it is possible to further reduce the memory footprint of the data structure. Consider a 3x3 block of obstacles in an otherwise empty map as in Fig. \ref{edge_map}. The center obstacle will never be the nearest neighbor in any ray casting query, because any such query would first intersect with one of the 8 surrounding obstacles. To exploit this, one can take the morphological edge map of the occupancy grid map prior to CDDT generation without loss of generality. During ray casting on the CDDT data structure, we check the occupancy grid to ensure that the query point itself is not an obstacle and return a distance of zero if so in order to prevent incorrect results when ray casting from the middle of obstacles removed during the morphological operation. In dense maps, this results in a significant reduction in both memory usage and construction time.

Additionally, consider a line of obstacles aligned along the X-axis. Every element in this line will be projected into a single zero point bin in the $\theta=0$ CDDT slice. However, the middle elements of the line will never result in a collision. Any ray cast from points on the line of obstacles will return early in the occupancy grid check, and any ray cast from non-overlapping points co-linear in the $\theta$ or $\theta+\pi$ directions will intersect either the first or last obstacle in the line. Therefore in the $\theta=0$ CDDT slice, one may discard the zero points corresponding to the middle elements without introducing error. This form of optimization is simple to compute in the cardinal directions, but non-trivial for arbitrary $\lfloor\theta\rceil$ not aligned with an axis. Rather than analytically determining which obstacles may be discarded, it is simpler to prune the data structure by ray casting from every possible state, discarding any unused zero point.

Pruning does increase pre-computation time, rendering it incompatible with incremental update. However, the reduction of memory usage is worthwhile for static maps. In addition to memory reduction, we find that pruning slightly improves runtime performance, likely as a result of improved caching characteristics and the reduced number of zero points. We refer to the pruned datastructure as PCDDT.

@@

The memory usage of the CDDT data structure is $O(n*\theta_{discretization}+|map|)$ where $n$ is the number of occupied pixels in the edge map. The occupancy grid map must be stored to check for overlaps between $(x,y)_{query}$ and obstacles prior to searching CDDT bins. For each edge pixel, a total of $\theta_{discretization}$ float values are stored in the CDDT bins. Due to the fact that we must sort each bin after CDDT construction, pre-computation time is at worst $O(n*\theta_{discretization} + S^2*\theta_{discretization}*log(S))$ for the same definition of n, where $S=norm(\text{map width},\text{map height})$. In practice it is closer to $O(n*\theta_{discretization} + S*\theta_{discretization})$ since each CDDT bin has a small number of elements on average, as evidenced by the high demonstrated compression ratio.

@@

The ray cast procedure has three general steps: projection into CDDT coordinate space, the search for nearby zero points, and the computation of distance given the nearest zero point. The first and last steps are simple arithmetic, and therefore are theoretically constant time. The second step involves searching for a value in a sorted array of zero points. As previously discussed in \ref{optimizations}, the number of zero points in each CDDT bin tends to be small and is bounded in map size. Thus, at worst the search operation requires $\log{(norm(\text{map width},\text{map height}))}$ which is a small constant value for a fixed map. Therefore, for a given map, our algorithm provides $O(1)$ query performance.

@@

We have implemented the proposed algorithm in the C++ programming language, as well as Bresenham’s Line, ray marching, and the LUT approach for comparison. Our source code$^1$ is available for use and analysis, and Python wrappers are also provided for easier usage. All benchmarks were performed on a desktop computer on a Intel Core i5-4590 CPU @ 3.30GHz with 16GB DDR3 ram @ 1333MHz, running Ubuntu 14.04.

\begin{figure}[h!]
\begin{center}
\begin{tabular}{ | m{8em} | m{2.3cm}| m{1.5cm} | } 
\hline
 \multicolumn{3}{|c|}{Basement Map, $\theta_{discretization}$: 108} \\
 \hline
 Method & Memory Usage & Init. Time \\
 \hline
 Bresenham's Line & 1.37 MB & 0.006 sec  \\
 Ray Marching & 5.49 MB & 0.16 sec  \\
 CDDT & 6.34 MB & 0.067 sec  \\
 PCDDT & 4.07 MB & 2.2 sec  \\
 Lookup Table & 296.63 MB & 15.3 sec  \\
\hline \hline
 \multicolumn{3}{|c|}{Synthetic Map, $\theta_{discretization}$: 108} \\
 \hline
 Method & Memory Usage & Init. Time \\
 \hline
 Bresenham's Line & 1 MB & 0.004 sec  \\
 Ray Marching  & 4 MB & 0.13 sec  \\
 CDDT & 2.71 MB & 0.03 sec  \\
 PCDDT & 1.66 MB & 0.74 sec  \\
 Lookup Table & 216 MB & 9.1 sec  \\
\hline
\end{tabular}
\end{center}
\caption{Construction time and memory footprint of each method. Ranges stored in lookup table with 16 bit integers, initialized with ray marching.}
\label{table:synthetic:init}
\end{figure}

\begin{figure}[ht]
\vspace{4pt}
\img{basement_grid_random_cropped.png}
% \begin{center}\includegraphics[width=248px,height=520px]{basement_benchmarks_cropped.png}\end{center}
\caption{Violin plots demonstrating histogram of completion time over a large number of queries for each ray cast method. Basement map. X axis shows time in milliseconds, and Y axis shows the number of queries that completed after that amount of time.}
\label{violin:basement:all}
\end{figure}

We evaluate algorithm performance in two synthetic benchmarks, using two different maps. The so called Synthetic map (Fig. \ref{synthetic_map}) was created with photoshop, whereas the basement map (Fig. \ref{sawtooth_to_cddt}) was created via a SLAM algorithm on the RACECAR platform\footnote{http://racecar.mit.edu} while navigating the Stata basement loop. The first benchmark computes a ray cast for each point in a uniformly spaced grid over the three dimensional state space. The second benchmark performs a large number of ray casts for randomly generated states.
